Network Visualizations in \MakeLowercase{\texttt{ggplot2}}

### 2015-10-08 inst/paper.Rnw
added bike example

<<readingdata, eval=TRUE, echo=FALSE>>=
library(ggplot2) # needs to be version ≥ 1.0.1.9003
library(ggnet) # install from github
data(blood, football, theme_elements, lesmis,madmen, mm_directed, email)

@@

@@

This network is plotted in figure \ref{fig.cap:email_ex}.  There are six distinct clusters in this network which almost perfectly correspond to the six different types of employee in this company: administration, engineering, executive, facilities, information technology, and security. Unfortunately, there is currently no legend to associate color with employment type. This will be remedied in the final version of our geometry. Additionally, the edges between employees in the same cluster are much darker than edges between employees in different clusters. This is due to the value of the \texttt{ealpha} aesthetic: more emails between two employees lead to darker edges.  The value is set to 0.1 in this example, so that 10 or more emails between 2 employees results in a completely opaque edge. This pattern of heavy communication between employees of the same type is fairly unsurprising. To make this visualization more interesting and informative, we facet the network by day: each panel in \ref{fig.cap:email_ex_facet} shows the different email networks associated with each day of the week.
%\begin{landscape}
\begin{figure}
\centering
<<email_ex_facet, fig.height=5, fig.width=8, fig.cap = 'All emails within a company over a two week period facetted by day of the month.',echo=TRUE, out.width='\\textwidth'>>=
ggplot(data = emailnet, aes(from_id = From, to_id = to)) +
  geom_net(aes(colour= CurrentEmploymentType), fiteach=TRUE,
           linewidth=0.5, alpha=0.2, size=2) +
  scale_colour_brewer(palette="Set1") + theme_net() +
  theme(legend.position="bottom") + facet_wrap(~day, nrow = 2)
\includegraphics[width=\textwidth]{figure/email_ex_facet-1.pdf}
\caption{\label{fig.cap:email_ex_facet} The same email network as in figure \ref{fig.cap:email_ex} facetted by day of the week.}
\end{figure}
%\end{landscape}
This plot is shown in figure \ref{fig.cap: email_ex_facet}. With the facetting, we can see that there are several days where one or more departments do not communicate with any of the other departments. There are only two days, 13 and 15, without any isolated department communications. Facetting is one of the major benefits of creating a geometry for networks in \texttt{ggplot2}.  Facetting quickly separates dense networks into separate subnetworks for easy visual comparison and  analyses.
  \subsection{\texttt{ggplot2} Theme Elements}
This example comes from the \texttt{theme()} help page in the \texttt{ggplot2} documentation \citep{ggplot}.  It is a directed network which shows the structure of the inheritance of theme options in the construction of a \texttt{ggplot2} plot.  There are 53 vertices and 36 edges in this network. Each vertex represents one possible theme option. There is an arrow from one theme option to another if the element represented by the \emph{to} vertex inherits its values from the \emph{from} vertex.  For example, the \texttt{axis.ticks.x} option inherits its value from the \texttt{axis.ticks} value, which inherits its value from the \texttt{line} option.  Thus, setting the \texttt{line} option to a value such as \texttt{element\_blank()} sets the entire inheritance tree to \texttt{element\_blank()}, and no lines appear anywhere on the plot background. Finally, we note that the vertices with no edges were incorporated into the plot by adding their labels to the edges data frame in both the  `from\_id' and `to\_id' columns before passing the edges data frame to \texttt{ggplot}.
<<theme_ex, echo=TRUE, fig.cap= 'Inheritance structure for \\texttt{theme} elements in \\texttt{ggplot2}.'>>=
ggplot(data = TEnet, aes(from_id = parent, to_id = child)) +
  geom_net(directed = TRUE, label = TRUE, vjust=-.5) +
  theme_net() + xlim(c(-0.05, 1.05))
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/theme_ex-1.pdf}
\caption{\label{fig.cap:theme_ex} Inheritance structure of \texttt{ggplot2} theme elements. This is a recreation of the graph found at \protect\url{http://docs.ggplot2.org/current/theme.html}.}
\end{figure}
\afterpage{\clearpage}

@@

>>>>

<<readingdata, eval=TRUE, echo=FALSE>>=
library(ggplot2) # needs to be version ≥ 1.0.1.9003
library(ggnet) # install from github
data(blood, football, theme_elements, lesmis,madmen, mm_directed, email,
     bikes)

@@

tripnet <- merge(bikes$trips, bikes$stations, by.x="Start.station", by.y="name", all=TRUE)
@

@@

This network is plotted in figure \ref{fig.cap:email_ex}.  There are six distinct clusters in this network which almost perfectly correspond to the six different types of employee in this company: administration, engineering, executive, facilities, information technology, and security. Unfortunately, there is currently no legend to associate color with employment type. This will be remedied in the final version of our geometry. Additionally, the edges between employees in the same cluster are much darker than edges between employees in different clusters. This is due to the value of the \texttt{ealpha} aesthetic: more emails between two employees lead to darker edges.  The value is set to 0.1 in this example, so that 10 or more emails between 2 employees results in a completely opaque edge. This pattern of heavy communication between employees of the same type is fairly unsurprising. To make this visualization more interesting and informative, we facet the network by day: each panel in \ref{fig.cap:email_ex_facet} shows the different email networks associated with each day of the week.
%\begin{landscape}
\begin{figure}
\centering
<<email_ex_facet, fig.height=5, fig.width=8, fig.cap = 'All emails within a company over a two week period facetted by day of the month.',echo=TRUE, out.width='\\textwidth'>>=
ggplot(data = emailnet, aes(from_id = From, to_id = to)) +
  geom_net(aes(colour= CurrentEmploymentType), fiteach=TRUE,
           linewidth=0.5, alpha=0.2, size=2) +
  scale_colour_brewer(palette="Set1") + theme_net() +
  theme(legend.position="bottom") + facet_wrap(~day, nrow = 2)
@
\includegraphics[width=\textwidth]{figure/email_ex_facet-1.pdf}
\caption{\label{fig.cap:email_ex_facet} The same email network as in figure \ref{fig.cap:email_ex} facetted by day of the week.}
\end{figure}
%\end{landscape}
This plot is shown in figure \ref{fig.cap: email_ex_facet}. With the facetting, we can see that there are several days where one or more departments do not communicate with any of the other departments. There are only two days, 13 and 15, without any isolated department communications. Facetting is one of the major benefits of creating a geometry for networks in \texttt{ggplot2}.  Facetting quickly separates dense networks into separate subnetworks for easy visual comparison and  analyses.
  \subsection{\texttt{ggplot2} Theme Elements}
This example comes from the \texttt{theme()} help page in the \texttt{ggplot2} documentation \citep{ggplot}.  It is a directed network which shows the structure of the inheritance of theme options in the construction of a \texttt{ggplot2} plot.  There are 53 vertices and 36 edges in this network. Each vertex represents one possible theme option. There is an arrow from one theme option to another if the element represented by the \emph{to} vertex inherits its values from the \emph{from} vertex.  For example, the \texttt{axis.ticks.x} option inherits its value from the \texttt{axis.ticks} value, which inherits its value from the \texttt{line} option.  Thus, setting the \texttt{line} option to a value such as \texttt{element\_blank()} sets the entire inheritance tree to \texttt{element\_blank()}, and no lines appear anywhere on the plot background. Finally, we note that the vertices with no edges were incorporated into the plot by adding their labels to the edges data frame in both the  `from\_id' and `to\_id' columns before passing the edges data frame to \texttt{ggplot}.
<<theme_ex, echo=TRUE, fig.cap= 'Inheritance structure for \\texttt{theme} elements in \\texttt{ggplot2}.'>>=
library(dplyr)
TEnet <- TEnet %>% group_by(parent) %>% mutate(degree=sqrt(10*n()+1))

ggplot(data = TEnet, aes(from_id = parent, to_id = child)) +
  geom_net(aes(fontsize=degree), directed = TRUE, label = TRUE, vjust=-.5,
           size=3) +
  theme_net() + xlim(c(-0.05, 1.05))
@
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figure/theme_ex-1.pdf}
\caption{\label{fig.cap:theme_ex} Inheritance structure of \texttt{ggplot2} theme elements. This is a recreation of the graph found at \protect\url{http://docs.ggplot2.org/current/theme.html}.}
\end{figure}
\afterpage{\clearpage}

@@

\subsection{Bikesharing in D.C.}
The data shows the second quartal trips in 2015 taken with bikes from the bike share company https://secure.capitalbikeshare.com/.
While this bikesharing company is located in the heart of Washington D.C.\ they do have a set of bike stations just outside of Washington in  Rockville, MD and north of it.
Each station is shown as a dot, lines between stations indicate that at least five trips were taken between these two stations; the wider the line, the more trips have been taken between stations. In order to reflect distance between stations, we use as an additional restriction that the fastest trip was at most ten minutes long.

<<bikes_ex, echo=TRUE, fig.cap="Bikesharing in Rockville", warning=FALSE, fig.keep='all'>>=
tripnet$Metro = FALSE
idx <- grep("Metro", tripnet$Start.station)
tripnet$Metro[idx] <- TRUE

ggplot(aes(from_id=Start.station, to_id=End.station), data=tripnet) +
  geom_net(aes(linewidth=n/15, colour=Metro), label=TRUE, vjust=-0.5) +
  theme_net() + xlim(c(-0.1, 1.1)) +
  scale_colour_manual(values=c("grey40", "darkorange"))
# compare to https://secure.capitalbikeshare.com/map/

ggplot(aes(from_id=Start.station, to_id=End.station),  data=tripnet) +
  geom_net(layout=NULL, label=TRUE, vjust=-0.5,
           aes(x=long, y=lat, linewidth=n/15, colour=Metro)) +
  theme_net() +
    scale_colour_manual(values=c("grey40", "darkorange"))

\begin{figure}
\centering
\includegraphics[width=.49\textwidth]{figure/bikes_ex-1.pdf}
\includegraphics[width=.49\textwidth]{figure/bikes_ex-2.pdf}
\caption{\label{fig:bikes} Network of bike trips using a Kamada-Kawai layout (left) and a geographically true representation (right). Metro stations are shown in orange. In the Kamada-Kawai layout based on the trips taken, metro stations take a much more central position than in the geographically true representation. }
\end{figure}
\afterpage{\clearpage}
